Expert-In-The-Loop Causal Discovery: Iterative Model Refinement Using Expert Knowledge

### 2025-02-09 paper/main.tex
Updates

In the DAG literature, the primary focus has been on developing automated
algorithms to learn causal structures from datasets. These efforts have led to
numerous causal discovery algorithms, such as constraint-based methods like PC
algorithm \citep{Spirtes2001} and Fast Causal Inference \citep{Spirtes2000}),
score-based methods such as Hill-Climb Search and Greedy Equivalence Search
\citep{Chickering2002}, and continuous optimization-based methods like NO TEARS
\citep{Zheng2018} and DAGMA \citep{Bello2022}.

@@

Despite significant progress in automated causal discovery, their adoption in
applied research has been limited. Researchers often prefer to manually
construct DAGs based on their domain expertise \citep{Tennant2020,
Petersen2021}. We attribute this preference to several challenges with existing
causal discovery algorithms in practical settings:

Figure~\ref{fig:intro} highlights some of these issues.   When manually
constructing models, it is important to test whether the resulting DAG
accurately represents the dataset. Based on this evaluation and domain
knowledge, we can make further modification to the model. One common approach
for assessing DAGs is to check whether the Conditional Independences (CIs)
implied by the DAG hold in the data \citep{Ankan2021}. If we find violations to
these CIs, we can use our domain knowledge to make appropriate changes to the
model.

>>>>

In the DAG literature, the primary focus has been on developing automated
algorithms to learn causal structures from datasets. These efforts have led to
numerous causal discovery algorithms, such as constraint-based methods like PC
algorithm \citep{Spirtes2001} and Fast Causal Inference \citep{Spirtes2000}),
score-based methods such as Hill-Climb Search and Greedy Equivalence Search
\citep{Chickering2002}, and continuous optimization-based methods like NO TEARS
\citep{Zheng2018} and DAGMA \citep{Bello2022}. Despite significant progress in
automated causal discovery, their adoption in applied research has been
limited. Researchers often prefer to manually construct DAGs based on their
domain expertise \citep{Tennant2020, Petersen2021}. We attribute this
preference to several challenges with existing causal discovery algorithms in
practical settings:

Figure~\ref{fig:intro} highlights some of these issues. When constructing DAGs
manually, it is important to test how consistent our DAG is to our dataset. One
of the methods for checking the consistency is by testing whether the implied
CIs of the DAG hold in the dataset or not \citep{Ankan2021}. As each pair of
variables that do not have an edge between them in the DAG implies a CI
statement, we can use a statistical test to check whether that statement holds
in our dataset or not. If we find any violations we can also modify our DAG
based on this and utilizing our domain knowledge to find a suitable orientation
for the edge between the pair of variables.

@@
